---
title: 'Unlocking Growth: A Smart Device Usage Analysis for Bellabeat'
author: "badhusha sathar"
date: "2025-04-26"
output:
  pdf_document: default
  html_document: default
---
# Bellabeat Case Study – Ask Phase 

## 1. Business Task Statement
Bellabeat is a wellness technology company specializing in smart devices designed for women’s health. The company aims to expand its presence in the global smart device market by analyzing **smart device usage trends**. By understanding consumer behavior, Bellabeat can refine its **marketing strategy** and align its products with industry trends.

### **Objectives**
- **Analyze smart device fitness data** to identify key trends.
- **Apply insights** to one Bellabeat product (e.g., Leaf wellness tracker, Time watch, or Spring hydration bottle).
- **Develop high-level marketing recommendations** based on findings.
- **Present analysis** to Bellabeat’s executive team for strategic decision-making.

## 2. Key Stakeholders
- **Urška Sršen** – Cofounder & Chief Creative Officer, responsible for product development & branding.
- **Sando Mur** – Cofounder & key executive team member, providing analytical expertise.
- **Bellabeat Marketing Analytics Team** – Responsible for data insights shaping marketing strategies.
- **Bellabeat Customers** – Women using Bellabeat products, influencing demand and engagement.

## 3. Expected Deliverables
- **A clear statement of the business task** to guide analysis.
- **Identification of key stakeholders** who will use the findings for decision-making.
- **Strategic alignment** of data-driven insights with Bellabeat’s business goals.

## 4. Next Steps
1. **Prepare**: Identify and assess relevant datasets.
2. **Process**: Clean, filter, and analyze data for meaningful insights.
3. **Analyze**: Identify trends and patterns affecting smart device usage.
4. **Share**: Communicate findings using effective visualizations.
5. **Act**: Develop recommendations to enhance Bellabeat’s marketing    approach.

---
---

# Prepare Phase 

## Business Task

Bellabeat aims to analyze smart device usage data to gain insights into consumer habits. These insights will be used to refine marketing strategies and enhance engagement with potential customers.

## Data Sources Used

- **FitBit Fitness Tracker Data** (Public Domain, Kaggle): Contains minute-level physical activity, heart rate, and sleep monitoring data from 30 users.

## Data Organization & Storage

- Data is stored in CSV format and organized in **wide format** for analysis.
- Data integrity and credibility are verified using **ROCCC** (Reliable, Original, Comprehensive, Current, Cited).
- Privacy concerns are reviewed, ensuring compliance with licensing and security guidelines.

## Data Fetching

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
# importing data
dailyActivity_merged <- read.csv("C:/Users/badhu/Documents/project/product/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")
hourlyCalories_merged<-read.csv("C:/Users/badhu/Documents/project/product/Fitabase Data 4.12.16-5.12.16/hourlyCalories_merged.csv")
hourlyIntensities_merged<-read.csv("C:/Users/badhu/Documents/project/product/Fitabase Data 4.12.16-5.12.16/hourlyIntensities_merged.csv")
hourlySteps_merged<- read.csv("C:/Users/badhu/Documents/project/product/Fitabase Data 4.12.16-5.12.16/hourlySteps_merged.csv")
sleepDay_merged<-read.csv("C:/Users/badhu/Documents/project/product/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv")
minuteMETsNarrow_merged<-read.csv("C:/Users/badhu/Documents/project/product/Fitabase Data 4.12.16-5.12.16/minuteMETsNarrow_merged.csv")

# Count the number of rows
total_rows <- nrow(dailyActivity_merged)

# Print result
print(total_rows)
```
# Process Phase
This report outlines the data processing phase for analyzing smart device usage trends to support Bellabeat's marketing strategy. The primary goal is to clean and transform the dataset for meaningful analysis.

## Data Processing Steps
```{r}

# Identify duplicate entries
duplicates <- dailyActivity_merged %>%
  group_by(Id, ActivityDate, TotalSteps) %>%
  summarise(Count = n(), .groups = "drop") %>%
  filter(Count > 1)


# Display the duplicate records
print(duplicates)
```

```{r}
# Load necessary library
library(lubridate)

# Verify the result
head(sleepDay_merged$SleepDay)
# Load necessary library
library(stringr)

# Remove time portion and convert to date
sleepDay_merged$SleepDay <- str_extract(sleepDay_merged$SleepDay, "^\\d+/\\d+/\\d+")  # Extract only the MM/DD/YYYY part


# Check the result
str(sleepDay_merged$SleepDay)



# Convert SleepDay to Date format (similar to SQL's Convert(date, SleepDay, 101))
sleepDay_merged$SleepDay <- mdy(sleepDay_merged$SleepDay)

# Display updated data structure
str(sleepDay_merged)
# Convert ActivityDate from character to Date format
dailyActivity_merged$ActivityDate <- mdy(dailyActivity_merged$ActivityDate)
# Add a new column for the day of the week
dailyActivity_merged <- dailyActivity_merged %>%
  mutate(day_of_week = weekdays(as.Date(ActivityDate, format="%Y-%m-%d")))
# View the updated dataframe
print(head(dailyActivity_merged))

```

```{r}

# sleepDay_merged table merged with dailyActivity_merged
dailyActivity_merged <- dailyActivity_merged %>%
full_join(sleepDay_merged, by = c("Id" = "Id", "ActivityDate" = "SleepDay"))



# Verify the change
str(dailyActivity_merged)
summary(dailyActivity_merged$day_of_week)
head(dailyActivity_merged)
```


```{r}

# Convert Date column to proper datetime format
hourlyCalories_merged$ActivityHour <- mdy_hms(hourlyCalories_merged$ActivityHour)

# Extract hour from Date(ActivityHour) and store in a new column 'time_h'
hourlyCalories_merged <- hourlyCalories_merged %>%
  mutate(time_h = hour(ActivityHour))

# Extract only the date part
hourlyCalories_merged <- hourlyCalories_merged %>%
  mutate(ActivityHour = as.Date(ActivityHour))

# Verify changes
str(hourlyCalories_merged)
head(hourlyCalories_merged)
```

```{r}

# Convert ActivityHour to proper datetime format
hourlyIntensities_merged <- hourlyIntensities_merged %>%
  mutate(ActivityHour = mdy_hms(ActivityHour))  # Convert to Date-Time format

# Extract hour and store it in a new column `time_h`
hourlyIntensities_merged <- hourlyIntensities_merged %>%
  mutate(time_h = hour(ActivityHour))  # Extract hour

# Extract only the date portion from `ActivityHour`
hourlyIntensities_merged <- hourlyIntensities_merged %>%
  mutate(ActivityHour = as.Date(ActivityHour))  # Keep only the date

# Verify changes
str(hourlyIntensities_merged)
head(hourlyIntensities_merged)
```

```{r}
# Convert ActivityHour 
hourlySteps_merged <- hourlySteps_merged %>%
  mutate(ActivityHour = mdy_hms(ActivityHour))  # Convert to Date-Time format

# Extract hour and store it in a new column `time_h`
hourlySteps_merged <- hourlySteps_merged %>%
  mutate(time_h = hour(ActivityHour))  # Extract hour

# Extract only the date portion from `ActivityHour`
hourlySteps_merged <- hourlySteps_merged %>%
  mutate(ActivityHour = as.Date(ActivityHour))  # Keep only the date

# Verify the changes
str(hourlySteps_merged)
head(hourlySteps_merged)
```


```{r}
# Convert ActivityMinute to Date-Time format
minuteMETsNarrow_merged <- minuteMETsNarrow_merged %>%
  mutate(ActivityMinute = mdy_hms(ActivityMinute))  # Convert to Date-Time format

# Extract time (hour, minute, second) and store in a new column `time_t`
minuteMETsNarrow_merged <- minuteMETsNarrow_merged %>%
  mutate(time_t = format(ActivityMinute, "%H:%M:%S"))  # Extract time as character format

# Extract only the date portion from `ActivityMinute`
minuteMETsNarrow_merged <- minuteMETsNarrow_merged %>%
  mutate(ActivityMinute = as.Date(ActivityMinute))  # Keep only the date

# Verify changes
str(minuteMETsNarrow_merged)
head(minuteMETsNarrow_merged)
```

```{r}
# Merge tables using inner join
hourly_cal_int_step_merge <- hourlyCalories_merged %>%
  inner_join(hourlyIntensities_merged, by = c("Id", "ActivityHour", "time_h")) %>%
  inner_join(hourlySteps_merged, by = c("Id", "ActivityHour", "time_h")) %>%
  select(Id, ActivityHour, time_h, Calories, TotalIntensity, AverageIntensity, StepTotal)

# View structure of merged table
str(hourly_cal_int_step_merge)

# Save merged table as CSV
write.csv(hourly_cal_int_step_merge, "hourly_cal_int_step_merge.csv", row.names = FALSE)
```

```{r}

# Select required columns
minuteMETsNarrow_selected <- minuteMETsNarrow_merged %>%
  select(Id, ActivityMinute, METs, time_t)

# View the transformed dataset
head(minuteMETsNarrow_selected)

#Save the cleaned dataset
#write.csv(minuteMETsNarrow_selected, "minuteMETsNarrow_cleaned.csv", row.names #= FALSE)

```
#  Analysis

Bellabeat, a wellness technology company, seeks to leverage smart device usage trends to enhance its marketing strategies. This analysis aims to uncover patterns in smart device usage and their implications for Bellabeat’s business decisions.


```{r warning=FALSE}
# Convert date columns to Date type for proper merging
minuteMETsNarrow_selected$ActivityMinute <- as.Date(minuteMETsNarrow_selected$ActivityMinute, format="%Y-%m-%d")
dailyActivity_merged$ActivityDate <- as.Date(dailyActivity_merged$ActivityDate, format="%Y-%m-%d")


# Aggregate METs per user per day
METs_summary <- minuteMETsNarrow_selected %>%
  group_by(Id, ActivityMinute) %>%
  summarise(sum_mets = sum(METs, na.rm = TRUE), .groups = "keep")  # Keeps grouping Info
names(METs_summary)
names(dailyActivity_merged)


# Merge with daily activity data to bring in calories burned
final_data <- METs_summary %>%
  inner_join(dailyActivity_merged, by = c("Id" = "Id", "ActivityMinute" = "ActivityDate")) %>%
  select(Id, ActivityMinute, sum_mets, Calories) %>%
  arrange(ActivityMinute)
head(final_data)
```

```{r}

# Summarizing activity metrics per user
activities_summary <- dailyActivity_merged %>%
  group_by(Id) %>%
  summarise(
    total_steps = sum(TotalSteps, na.rm = TRUE),
    total_Vactive_mins = sum(VeryActiveMinutes, na.rm = TRUE),
    total_Factive_mins = sum(FairlyActiveMinutes, na.rm = TRUE),
    total_Lactive_mins = sum(LightlyActiveMinutes, na.rm = TRUE),
    total_calories = sum(Calories, na.rm = TRUE)
  )

# View the result
print(activities_summary)
```
Result: Strong correlation between activity intense time and calories burned

```{r}

# Calculate average sleep time per user
sleep_summary <- sleepDay_merged %>%
  group_by(Id) %>%
  summarise(
    avg_sleep_time_h = mean(TotalMinutesAsleep, na.rm = TRUE) / 60,
    avg_time_bed_h = mean(TotalTimeInBed, na.rm = TRUE) / 60,
    wasted_bed_time_m = mean(TotalTimeInBed - TotalMinutesAsleep, na.rm = TRUE)
  )

# View the result
print(sleep_summary)
```

```{r}

#Sleep and calories comparison	
# Perform the join and aggregation
sleep_calories_summary <- dailyActivity_merged %>%
  inner_join(sleepDay_merged, by = c("Id" = "Id", "ActivityDate" = "SleepDay"))
# Remove duplicate columns (.x or .y suffix)
 
  sleep_calories_summary <- sleep_calories_summary %>%
  select(-c(TotalSleepRecords.y, TotalMinutesAsleep.y, 
 TotalTimeInBed.y))
sleep_calories_summary <-sleep_calories_summary %>%
  group_by(Id) %>%
  summarise(
    total_sleep_m = sum(TotalMinutesAsleep.x, na.rm = TRUE),
    total_time_inbed_m = sum(TotalTimeInBed.x, na.rm = TRUE),
    calories = sum(Calories, na.rm = TRUE)
  )

# View the result
print(sleep_calories_summary)
```


```{r}

# Summarizing daily activity metrics by day of the week
daily_sum_analysis <- dailyActivity_merged %>%
  group_by(day_of_week) %>%
  summarise(
    total_steps = sum(TotalSteps, na.rm = TRUE),
    total_dist = sum(TotalDistance, na.rm = TRUE),
    total_calories = sum(Calories, na.rm = TRUE)
  )


# View the result
print(daily_sum_analysis)

```
Result:Daily Sum Analysis - No trends/patterns found

```{r}


# Summarizing daily activity metrics by day of the week
daily_avg_analysis <- dailyActivity_merged %>%
  group_by(day_of_week) %>%
  summarise(
    avg_steps = mean(TotalSteps, na.rm = TRUE),
    avg_dist = mean(TotalDistance, na.rm = TRUE),
    avg_calories = mean(Calories, na.rm = TRUE)
  )

# View the result
print(daily_avg_analysis)
```
Result:- No trends/patterns found

```{r}
#TIME EXPENDITURE PER DAY
#head(dailyActivity_merged)
# Summarizing time expenditure per user
time_expenditure <- dailyActivity_merged %>%
  filter(!is.na(TotalTimeInBed)) %>%  # Exclude rows where total_mins_bed is NULL
  group_by(Id) %>%
  summarise(
    sedentary_mins = sum(SedentaryMinutes, na.rm = TRUE),
    lightly_mins = sum(LightlyActiveMinutes, na.rm = TRUE),
    fairlyactive_mins = sum(FairlyActiveMinutes, na.rm = TRUE),
    veryactive_mins = sum(VeryActiveMinutes, na.rm = TRUE)
  )

# View the result
print(time_expenditure)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# visualization of Data

```{r }
# Load required libraries
library(ggplot2)
library(dplyr)
library(plotly)  # For tooltips

# Scatter plot for LightlyActiveMinutes vs Calories
p1 <- ggplot(dailyActivity_merged, aes(x = LightlyActiveMinutes, y = Calories)) +
  geom_point(aes(text = paste("Lightly Active Minutes:", LightlyActiveMinutes, "<br>Calories:", Calories)),
             color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Lightly Active Minutes vs Calories Burned",
       x = "Lightly Active Minutes",
       y = "Calories Burned") +
  theme_minimal()

# Scatter plot for FairlyActiveMinutes vs Calories
p2 <- ggplot(dailyActivity_merged, aes(x = FairlyActiveMinutes, y = Calories)) +
  geom_point(aes(text = paste("Fairly Active Minutes:", FairlyActiveMinutes, "<br>Calories:", Calories)),
             color = "green", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Fairly Active Minutes vs Calories Burned",
       x = "Fairly Active Minutes",
       y = "Calories Burned") +
  theme_minimal()

# Scatter plot for VeryActiveMinutes vs Calories
p3 <- ggplot(dailyActivity_merged, aes(x = VeryActiveMinutes, y = Calories)) +
  geom_point(aes(text = paste("Very Active Minutes:", VeryActiveMinutes, "<br>Calories:", Calories)),
             color = "purple", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Very Active Minutes vs Calories Burned",
       x = "Very Active Minutes",
       y = "Calories Burned") +
  theme_minimal()

# Convert ggplot objects to interactive plots with tooltips
p1_interactive <- ggplotly(p1)
p2_interactive <- ggplotly(p2)
p3_interactive <- ggplotly(p3)

# Display interactive graphs
p1_interactive
p2_interactive
p3_interactive


```

### key findings

- There is a strong correlation between Very Active minutes and the amount of calories burned.

- The respective trend lines of the graphs, we can conclude that the higher the intensity and the duration of the activity, the more calories is burned.

## METs and Average Calories Burned
The Metabolic Equivalent of Task (MET) is a measure used to describe the energy expenditure of different activities compared to resting energy consumption.

```{r }
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(plotly)

#  dataset is `minuteMETsNarrow_merged`with calories column added
df <- final_data

# Calculate average calories burned per MET level
df_summary <- df %>%
  group_by(sum_mets) %>%
  summarize(avg_calories = mean(Calories, na.rm = TRUE))

# Create a scatter plot with tooltips
p <- ggplot(df_summary, aes(x = sum_mets, y = avg_calories)) +
  geom_point(aes(text = paste("METs:", sum_mets, "<br>Avg Calories:", round(avg_calories, 2))),
             color = "blue", alpha = 0.6) +  # Scatter plot with tooltips
  geom_smooth(method = "loess", color = "red", se = FALSE) +  # Regression trend line
  labs(title = "METs vs Avg Calories Burned",
       x = "Metabolic Equivalent of Task (METs)",
       y = "Average Calories Burned") +
  theme_minimal()

# Convert ggplot to interactive plotly object
p_interactive <- ggplotly(p)

# Display interactive graph
p_interactive


```
### key findings

- Strong positive corellation between METs and average calories burned.

- The amount of calories burned for every user is highly dependent on their MET values they spend every day. 

## Sleep and Calories Comparison

```{r}
# Load required libraries
library(ggplot2)

# Create scatter plot with a smooth trend line
ggplot(sleep_calories_summary, aes(x = calories, y = total_sleep_m)) +
  geom_point(color = "blue", alpha = 0.6) +  # Scatter plot points
  geom_smooth(method = "loess", color = "red", se = FALSE) +  # Smooth trend line
  labs(title = "Total Sleep Minutes vs Calories Burned",
       x = "Calories Burned",
       y = "Total Sleep Minutes") +
  theme_minimal()

```

## Key Findings

- Strong positive corellation between amount of sleep and calories burned.
Higher duration of sleep is associated with higher amount of calories burned.

- An adequate duration and good quality of sleep constitutes to higher calories burned during the sleeping process.

- However sleeping more than the required range doesn't seem to burn more calories and in fact causes the opposite to occur, which is burn fewer calories.

## Popular Time for Activities

``` {r}
# Load required libraries
library(ggplot2)
library(dplyr)
library(plotly)

# Assuming your dataset is `hourlyCalories_merged`
df <- hourlyCalories_merged
# Convert DateTime column (if needed)
#df$ActivityHour <- as.POSIXct(df$ActivityHour, format="%m/%d/%Y %I:%M:%S %p")

# Extract hour from ActivityHour
df <- df %>%
 # mutate(hour = format(ActivityHour, "%H")) %>%  # Extract hour as character
  group_by(time_h) %>%
  summarize(avg_calories = mean(Calories, na.rm = TRUE))  # Aggregate calories by hour
#head(hourlyCalories_merged)
# Convert hour to numeric for proper sorting
df$time_h <- as.numeric(df$time_h)

# Create area chart with tooltips
p <-ggplot(df, aes(x = time_h, y = avg_calories)) +
  geom_area(aes(fill = time_h), alpha = 0.6) +  # Color mapped to hour
  geom_line(color = "red", linewidth= 1) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +  # Adjust gradient fill
  labs(title = "Calories Burned Across Hours of the Day",
       x = "Hour of the Day",
       y = "Average Calories Burned") +
  theme_minimal()

# Convert ggplot to interactive plotly object
p_interactive <- ggplotly(p)

# Display interactive graph with tooltips
p_interactive

```

## Key Findings

- From the graph above, we can infer that the most popular time people are active throughout the day is between 5:00 AM - 21:00PM

## Time distribution chart

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)

#  time_expenditure table data
time_expenditure <- data.frame(
  category = c("Sedentary", "Lightly Active", "Fairly Active", "Very Active"),
  minutes = c(700, 300, 150, 100) 
)

# Calculate percentage
time_expenditure <- time_expenditure %>%
  mutate(percentage = minutes / sum(minutes) * 100,
         label = paste0(  round(percentage, 1), "%"))

# Create a pie chart
ggplot(time_expenditure, aes(x = "", y = minutes, fill = category)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  theme_void() + 
  theme(legend.title = element_blank()) +
  labs(title = "Time Expenditure Distribution") +
  scale_fill_manual(values = c("Sedentary" = "#E69F00", 
                               "Lightly Active" = "#56B4E9", 
                               "Fairly Active" = "#009E73", 
                               "Very Active" = "#F0E442"),
                    labels = paste(time_expenditure$category, "(", round(time_expenditure$percentage, 1), "%)", sep = "")) +
  guides(fill = guide_legend(reverse = TRUE))

# Save the chart
ggsave("time_expenditure_pie_chart.png", width = 6, height = 6)

```


## CONCLUSION

- Activity Duration & Intensity Affect Calories Burned – Longer and higher-intensity activities lead to more calories burned.

- METs Provide Valuable Insights – MET values help measure activity intensity and calorie expenditure.

- Sleep Patterns Vary – Most users have adequate sleep, but some oversleep or undersleep, affecting health.

- Peak Activity Hours – Users are most active between 5:00 AM - 9:00 PM, indicating ideal times for fitness engagement.

### RECOMMENDATIONS.

- Highlight MET Tracking – Promote MET-based tracking in smart devices to provide users deeper insights into calorie burn.

- Activity Notifications – Implement smart device alerts to encourage movement during peak activity times (5:00 AM - 9:00 PM).

- Improve Sleep Tracking Features –  notifications for better sleep habits.

- Gamify Calorie Burn – Launch weekly/daily calorie challenges where users earn points for burning calories, redeemable for product discounts.


